Avoiding tipping points in ecosystem based fisheries management: a non-parametric Bayesian approach

### 2014-05-12 manuscripts/manuscript.Rmd

---
layout: 12pt,review
title: "Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach"
author: 
  - name: "Carl Boettiger"
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Marc Mangel
    affiliation: cstar
  - name: Stephan Munch
    affiliation: noaa
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: noaa
    address: | 
      Southwest Fisheries Science Center, 
      National Oceanic and Atmospheric Administration, 
      110 Shaffer Road, Santa Cruz, CA 95060, USA

abstract: |
          Model uncertainty and limited data coverage are fundamental
          challenges to robust ecosystem management.  These challenges
          are acutely highlighted by concerns that many ecological systems
          may contain tipping points.  Before a collapse, we do not know
          where the tipping points lie, if the exist at all.  Hence,
          we know neither a complete model of the system dynamics nor
          do we have access to data in some large region of state-space
          where such a tipping point might exist.  These two sources of
          uncertainty frustrate state-of-the-art parametric approaches
          to decision theory and optimal control.  Here we illustrate
          how a non-parametric approach using a Gaussian process prior
          provides a more flexible representation of this inherent
          uncertainty. We can then embed the Gaussian process in a
          stochastic dynamic programming framework in order to make
          robust management predictions under both model uncertainty and
          limited data. We conduct a set of simulations to evaluate this
          approach as compared with the standard approach of using model
          selection to choose from a set of candidate models.  In keeping
          with previous results, we find that model selection erroneously
          favors models without tipping points -- leading to harvest policies
          that guarantee extinction. The Gaussian process approach
          performs nearly as well as the true model and significantly
          outperforms standard appraoches.

@@

Decision making under uncertainty is a ubiquitous challenge of
natural resource management and conservation.  Ecological dynamics are
frequently complex and difficult to measure, making uncertainty in our
understanding a persistent challenge to effective management.
Decision-theoretic approaches provide a framework to determine the
best sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified [@Fischer2009].  The sudden
collapse of fisheries and other ecosystems has increasingly emphasized
the difficulties of formulating even qualitatively correct models of
the underlying processes.

@@

@@

We consider three candidate parametric models of the stock-recruitment
dynamics: The Ricker model, the Allen model [@Allen2005], the Myers
model [@Myers1995]. The familiar Ricker model involves two parameters, corresponding
to a growth rate and a carrying capacity, and cannot support alternative
stable state dynamics (though as growth rate increases it exhibits a
periodic attractor that proceeds through period-doubling into chaos. We
will generally focus on dynamics below the chaotic threshold for
the purposes of this analysis.) The Allen model resembles the Ricker
dynamics with an added Allee effect parameter [@Courchamp2008], below
which the population cannot persist.  The Myers model also has three
parameters and contains an Allee threshold, but has compensatory rather
than over-compensatory density dependence (resembling a Beverton-Holt
curve rather than a Ricker curve at high densities.)  
We assume multiplicative log-normal noise perturbs the growth predicted 
by the each of the deterministic model skeletons described above. This 
introduces one additional parameter $\sigma$ that must be estimated by each
model.

<!-- equations just in appendix? -->
Because of our interest in management performance in the presence of
tipping points, all of our simulations are based on the Allen model.
The Allen model is thus structurally correct and is expected to provide
a best-case scenario when the 'true' dynamics are known.
The Ricker model is a reasonable approximation of these dynamics far from the Allee
threshold (but lacks threshold dynamics), while the Myers model shares
the essential feature of a threshold but differs in the structure.

@@

We inferred posterior distributions for the parameters of each model
in a Bayesian context using Gibbs sampling (implemented in R [@RTeam]
using jags, [@R2jags]).  We choose uninformative uniform priors for all
parameters (See Appendix, Figures and tables, and the R code
provided). One-step-ahead predictions of these model fits are shown in
Figure 1.  Each chain was tested for Gelman-Rubin convergence and results were robust
to longer runs.  For each simulation we also applied several commonly
used model selection criteria (AIC, BIC, DIC, see @Burnham2002)
to identify the best fitting model.

@@

We use a Metropolis-Hastings Markov Chain Monte Carlo to infer posterior
distributions of the two parameters of the GP (Figure S13, code in
appendix), under weakly informative Gaussian priors (see parameters in
table S5). As the posterior distributions differ substantially from the
priors (Figure S13), we can be assured that most of the information in
the posterior comes from the data rather than the prior belief.

@@

It is reasonably straight-forward to derive the harvest policy from the
estimated BP by inserting it into a Stochastic Dynamic Programming (SDP)
algorithm.  Recall that unlike the parametric models the Gaussian process
with fixed parameters already predicts a distribution of curves rather
than a single curve. We must first integrate over this distribution of
curves given a sampling of parameter values drawn from the posterior
distribution of the two GP parameters, before integrating over the
posterior of those parameters themselves.

@@

```{r Figureb_posteriors, dependson=c("assemble-models", "par-fns"), include=TRUE, echo=FALSE, fig.cap="Points show the training data of stock-size over time.  Curves show the posterior step-ahead predictions based on each of the estimated models.", fig.width=8, fig.height=6}
```

@@

```{r out_of_sample_predictions, include=TRUE, echo=FALSE, fig.width=8, fig.height=6, fig.cap="Out of sample predictions of the dynamics under each model.  Points show the stock size simulated by the true model.  Overlay shows the range of states predicted by each model, based on the state observed in the previous time step. The Ricker model always predicts population growth, while the actual population shrinks in each step as the initial condition falls below the Allee threshold of the underlying model (Allen).  Note that the GP is both more pessimistic and more uncertain about the future state than the parametric models, while the realized state often falls outside of the expected range forecast by the structurally incorrect Myers and Ricker models." }
```

@@

```{r Figure3, dependson=c("sim"), fig.cap="Gaussian process inference outperforms parametric estimates. Shown are 100 replicate simulations of the stock dynamics (eq 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model.",  include=TRUE, echo=FALSE}
```

@@

In any modeling effort, models must be chosen for the task at hand.
Though simple mechanistically motivated models offer the greatest
potential to increase our basic understanding of ecological processes
[@Cuddington2013; @Geritz2012], such models can be not only inaccurate but
misleading when relied upon in a quantitative decision making framework.
In this paper we have tackled two aspects of uncertainty that are both
common to many ecological decision-making problems and fundamentally
challenging to existing approaches which largely rely on parametric
models:

@@

<!--
### Higher dimensions 
-->

@@

<!--
### Real-time learning
-->

@@

This work was partially supported by NOAA-IAM grant to SM administered through the Center for Stock Assessment
Research, a partnership between the University of California Santa Cruz
and the Fisheries Ecology Division, Southwest Fisheries Science Center,
Santa Cruz, CA and by NSF grant EF-0924195 to MM and NSF grant DBI-1306697
to CB.

\appendix

Model definitions and estimation
=================================

@@

The Ricker model is given by

(@rickereq) $$X_{t+1} = Z_t X_t e^{r \left(1 - \frac{S_t}{K} \right) } $$

where $Z_t$ is log-normal noise of mean unity and log standard deviation $\sigma$, representing the stochastic growth, $X_t$ the stock size at time $t$, $S_t$ the escapement (unharvested population that will recruit in the following year, $S_t = X_t - h_t$).  We place uniform priors on the growth rate $r$, carrying capacity $K$, and log-normal standard deviation parameter $\sigma$, over ranges given in Table 1. Posteriors are inferred by Gibbs sampling using Jags [@R2jags] (see provided code).

@@

```{r, echo=FALSE, include=TRUE, fig.height=6, fig.cap="Traces from the MCMC estimate of the Ricker model", fig.width=6}
plot_ricker_traces
```

@@

```{r echo=FALSE, results = "asis", include = TRUE, cache=FALSE}
xtable::xtable(ricker_priors_xtable,
  caption = "parameterization range for the uniform priors in the Ricker model")
```

@@

The Myers model [@Myers1995] is given by

<!--Eq S2: Myers model -->
(@myerseq) $$ X_{t+1} = Z_t \frac{r S_t^{\theta}}{1 - \frac{S_t^\theta}{K}}$$

where $Z_t$ is log-normal noise of mean unity and log standard deviation $\sigma$, representing the stochastic growth, $X_t$ the stock size at time $t$, $S_t$ the escapement (unharvested population that will recruit in the following year, $S_t = X_t - h_t$).  We place uniform priors on the growth rate $r$, carrying capacity $K$, $\theta$ controls the strength of the nonlinearity, exhibiting an allee effect for $\theta \geq 2$, and log-normal standard deviation parameter $\sigma$, over ranges given in Table 1. Posteriors are inferred by Gibbs sampling using Jags [@R2jags] (see code provided).

@@

```{r, echo=FALSE, include=TRUE, fig.height=6, fig.cap="Traces from the MCMC estimate of the Myers model", fig.width=6}
plot_myers_traces
```

@@

```{r echo=FALSE, results="asis", include = TRUE, cache=FALSE}
xtable::xtable(myers_priors_xtable,
           caption = "parameterization range for the uniform priors in the Myers model")
```

@@

The Allen model [@Allen2005] is given by

(@alleneq) $$f(S_t) = S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(S_t - C\right)} $$

where $Z_t$ is log-normal noise of mean unity and log standard deviation $\sigma$, representing the stochastic growth, $X_t$ the stock size at time $t$, $S_t$ the escapement (unharvested population that will recruit in the following year, $S_t = X_t - h_t$).  We place uniform priors on the growth rate $r$, carrying capacity $K$, allee threshold $C$, and log-normal standard deviation parameter $\sigma$, over ranges given in Table 1. Posteriors are inferred by Gibbs sampling using Jags [@R2jags] (see code provided).

@@

```{r, echo=FALSE, include=TRUE, fig.height=6, fig.cap="Traces from the MCMC estimate of the Allen model", fig.width=6}
plot_allen_traces
```

@@

```{r echo=FALSE, results = "asis", include=TRUE, cache=FALSE}
xtable::xtable(allen_priors_xtable,
  caption = "parameterization range for the uniform priors in the Allen model")
```

@@

}

@@

```{r echo=FALSE, include=TRUE, fig.cap = "Traces from the MCMC estimates of the GP model"}
gp_assessment_plots$traces_plot
```

@@

@@

Where $Z_t$ is multiplicative noise function with mean 1, representing
stochastic growth. We will consider log-normal noise with shape parameter
$\sigma_g$.

@@

For simplicity we assume profit is simply linear in the realized harvest (only
enforcing the restriction that harvest can not exceed available stock), $\Pi(h,x) = min(h,x)$.

### Pseudocode for the Bellman iteration

```r
# compute the value for each possible harvest
for(h in 1:length(h_grid)){
  V1[h] = delta * F[[h]] %*% V +  profit(x_grid, h_grid[h]) 
}
# find havest h that gives the maximum value
for(j in 1:gridsize){
  value = max(V1[j,], na.rm = T)  # each col is a different h, max over these
  index = which.max(V1[j,])       # store index so we can recover h's 
  output[,j] = c(value, index)    # returns both profit value & index of optimal h.  
}
# Sets V[t+1] = max_h(V[t]) at each possible state value, x
V = out[1,]                        # The new value-to-go
D[,OptTime-time+1] = out[2,]       # The index positions
```

@@

All code used in producing this analysis can be found in the R package accompanying the paper, available from [http://github.com/cboettig/nonparametric-bayes]. The code for all results shown here is dynamically embedded into the manuscript using Knitr, [@knitr].

>>>>

---
layout: 12pt,review
linenumbers: true
title: "Avoiding tipping points in ecosystem based fisheries management: a non-parametric Bayesian approach"
author: 
  - name: "Carl Boettiger"
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Marc Mangel
    affiliation: cstar
  - name: Stephan Munch
    affiliation: noaa
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: noaa
    address: | 
      Southwest Fisheries Science Center, 
      National Oceanic and Atmospheric Administration, 
      110 Shaffer Road, Santa Cruz, CA 95060, USA

abstract: |
          Model uncertainty and limited data coverage are fundamental
          challenges to robust ecosystem management.  These challenges
          are acutely highlighted by concerns that many ecological systems
          may contain tipping points.  Before a collapse, we do not know
          where the tipping points lie, if they exist at all.  Hence,
          we know neither a complete model of the system dynamics nor
          do we have access to data in some large region of state-space
          where such a tipping point might exist.  These two sources of
          uncertainty frustrate state-of-the-art parametric approaches
          to decision theory and optimal control.  Here we illustrate
          how a non-parametric approach using a Gaussian process prior
          provides a more flexible representation of this inherent
          uncertainty. We can then embed the Gaussian process in a
          stochastic dynamic programming framework in order to make
          robust management predictions under both model uncertainty and
          limited data. We conduct a set of simulations to evaluate this
          approach as compared with the standard approach of using model
          selection to choose from a set of candidate models.  In keeping
          with previous results, we find that model selection erroneously
          favors models without tipping points -- leading to harvest policies
          that guarantee extinction. The Gaussian process approach
          performs nearly as well as the true model and significantly
          outperforms standard appraoches. We illustrate this using examples
          of simulated single-species dynamics, where the standard model
          selection approach should be most effective, and find that it
          still fails to account for uncertainty appropriately and leads
          to population crashes, while management based on the Gaussian 
          process approach does not. We discuss opportunities to adapt
          this approach to multiple species models in further work.

@@

Decision making under uncertainty is a ubiquitous challenge of
natural resource management and conservation.  Ecological dynamics
are frequently complex and difficult to measure, making uncertainty
in our understanding a persistent challenge to effective management.
Decision-theoretic approaches provide a framework to determine the
best sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified [@Fischer2009].  Concerns
about the potential sudden collapse of fisheries and other ecosystems
[@Worm2006, @Scheffer2009] has increasingly emphasized the difficulties of
formulating even qualitatively correct models of the underlying processes.

@@

We focus on a class of one-dimensional models of population dynamics as a
'best-case' scenario for the parametric model performance relative to
the Gaussian process.   We consider this a best-case scenario for the
following reasons: first, the underlying dynamics are always simuated
from one the parametric models in the set, whereas in reality we never
have the "true" model among our candidates.  By choosing one-dimensional
models with few parameters, we limit the chance that poor performance
will be due to our inability to estimate parameters well, something
that becomes a more severe problem for higher-dimensional parametric
models. Further, the parametric models we consider are those most commonly
used in modeling stock-recruitment dynamics or to model sudden transitions
between alternative stable states.

@@

We consider three candidate parametric models of the stock-recruitment
dynamics: The Ricker model, the Allen model [@Allen2005], the Myers
model [@Myers1995], equations~\ref{ricker}-\ref{myers}, which we parameterize as follows:

We consider three parametric models that relate population size at time $t+1$, $X_{t+1}$, to population size at time $t$, $X_t$  and harvest at time $t$, $h_t$ by

$$ X_{t+1}= Z_t f(S_t,p) $$

Where Z(t) is log-normally distributed process stochasticity [@Reed1979] and $p$ is a vector of parameters, and $S_t$ the escapement [@Reed1979]

$$S_t = X_t - h_t$$

or the Ricker model,

(@ricker) $$ f(S(t)|p) = S_t e^{r \left(1 - \frac{S_t}{K} \right) } $$

where $K$ is the carrying capacity and $r$ the growth rate.  For the Allen model,

(@allen) $$ f(S(t)|p) = S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(S_t - X_C\right)} $$

Where $r$ and $K$ are as before, and $X_c$ denotes the location of the unstable (tipping) point.  For the Myers model,

(@myers) $$ f(S(t) | p)  = \frac{r S_t^{\theta}}{1 - \frac{S_t^\theta}{K}}$$

Where $r$ and $K$ as before, and $\theta$ the nonlinearity of the Myers model (corresponding to Beverton-Holt dynamics for $\theta$ = 1 and an Allee effect with alternative stable states exist at $\theta > 2$).

The familiar Ricker model involves two parameters, corresponding
to a growth rate and a carrying capacity, and cannot support alternative
stable growth rate dynamics. The Allen model resembles the Ricker
dynamics with an added Allee effect parameter [@Courchamp2008], below
which the population cannot persist.  The Myers model also has three
parameters and contains an Allee threshold, but has compensatory rather
than over-compensatory density dependence (resembling a Beverton-Holt
curve rather than a Ricker curve at high densities.)  
We assume multiplicative log-normal noise perturbs the growth predicted 
by the each of the deterministic model skeletons described above. This 
introduces one additional parameter $\sigma$ that must be estimated by each
model.

Because of our interest in management performance in the presence of
tipping points, all of our simulations are based on the Allen model.
The Allen model is thus structurally correct and is expected to provide
a best-case scenario when the 'true' dynamics are known.
The Ricker model is a reasonable approximation of these dynamics far from the Allee
threshold (but lacks threshold dynamics), while the Myers model shares
the essential feature of a threshold but differs in the structure.

@@

We inferred posterior distributions for the parameters of each model
in a Bayesian context using Gibbs sampling (@Gelman2003 implemented in R [@RTeam]
using jags, [@R2jags]).  We choose uninformative uniform priors for all
parameters (See Appendix, Figures and tables, and the R code
provided). We show One-step-ahead predictions of these model fits in
Figure 1.  Each chain was tested for Gelman-Rubin convergence and results were robust
to longer runs.  For each simulation we also applied several commonly
used model selection criteria (AIC, BIC, DIC, see @Burnham2002)
to identify the best fitting model.

@@

We use a Metropolis-Hastings Markov Chain Monte Carlo (@Gelman2003) to infer posterior
distributions of the two parameters of the GP (Figure S13, code in
appendix), under weakly informative Gaussian priors (see parameters in
table S5). As the posterior distributions differ substantially from the
priors (Figure S13), we can be assured that most of the information in
the posterior comes from the data rather than the prior belief.

@@

It is reasonably straight-forward to derive the harvest policy from the
estimated BP by inserting it into a SDP
algorithm.  Recall that unlike the parametric models the Gaussian process
with fixed parameters already predicts a distribution of curves rather
than a single curve. We must first integrate over this distribution of
curves given a sampling of parameter values drawn from the posterior
distribution of the two GP parameters, before integrating over the
posterior of those parameters themselves.

@@

```{r Figureb_posteriors, dependson=c("assemble-models", "par-fns"), include=TRUE, echo=FALSE, fig.cap="Points show the training data of stock-size over time.  Curves show the posterior step-ahead predictions based on each of the estimated models. Observe that all models are fitting the data reasonably well.", fig.width=8, fig.height=6}
```

@@

```{r out_of_sample_predictions, include=TRUE, echo=FALSE, fig.width=8, fig.height=6, fig.cap="Beyond the tipping point (outside the range of the training data), the true dynamics fall outside the uncertainty bands of the wrong parametric models, but inside the uncertainty predicted by the GP. Points show the stock size simulated by the true model.  Overlay shows the range of states predicted by each model, based on the state observed in the previous time step. The Ricker model always (wrongly) predicts positive population growth, while the actual population shrinks in each step as the initial condition falls below the Allee threshold of the underlying model (Allen).  Note that the GP is both more pessimistic and more uncertain about the future state than the parametric models." }
```

@@

```{r Figure3, dependson=c("sim"), fig.cap="Gaussian process inference outperforms parametric estimates. Shown are 100 replicate simulations of the stock dynamics (eq 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model.",  include=TRUE, echo=FALSE}

ggplot(sims_data) + 
  geom_line(aes(time, fishstock, group=interaction(reps,method), color=method), alpha=.1) +
  scale_colour_manual(values=colorkey, guide = guide_legend(override.aes = list(alpha = 1))) + 
  facet_wrap(~method) + guides(legend.position="none")

```

@@

Though simple mechanistically motivated models offer the greatest
potential to increase our basic understanding of ecological processes
[@Cuddington2013; @Geritz2012], such models can be not only inaccurate but
misleading when relied upon in a quantitative decision making framework.
In this paper we have tackled two aspects of uncertainty that are both
common to many ecological decision-making problems and fundamentally
challenging to existing approaches which largely rely on parametric
models:

@@

@@

@@

This work was partially supported by NOAA-IAM grant to SM and Alec McCall and administered
through the Center for Stock Assessment Research, a partnership between
the University of California Santa Cruz and the Fisheries Ecology
Division, Southwest Fisheries Science Center, Santa Cruz, CA and by NSF
grant EF-0924195 to MM and NSF grant DBI-1306697 to CB.

\appendix
\renewcommand\thefigure{\thesection.\arabic{figure}}

Appendix: Model definitions and estimation
===========================================

@@

@@

```{r, echo=FALSE, include=TRUE, fig.height=6, fig.cap="Traces from the MCMC estimates of the Ricker model show reasonable mixing (no trend) and sampling rejection rate (no piecewise jumps)", fig.width=6}
plot_ricker_traces
```

@@

```{r echo=FALSE, results = "asis", include = TRUE, cache=FALSE}
xtable::xtable(ricker_priors_xtable,
  caption = "Parameterization range for the uniform priors in the Ricker model")
```

@@

@@

```{r, echo=FALSE, include=TRUE, fig.height=6, fig.cap="Traces from the MCMC estimates of the Myers model show reasonable mixing (no trend) and sampling rejection rate (no piecewise jumps)", fig.width=6}
plot_myers_traces
```

@@

```{r echo=FALSE, results="asis", include = TRUE, cache=FALSE}
xtable::xtable(myers_priors_xtable,
           caption = "Parameterization range for the uniform priors in the Myers model")
```

@@

@@

```{r, echo=FALSE, include=TRUE, fig.height=6, fig.cap="Traces from the MCMC estimates of the Allen model show reasonable mixing (no trend) and sampling rejection rate (no piecewise jumps)", fig.width=6}
plot_allen_traces
```

@@

```{r echo=FALSE, results = "asis", include=TRUE, cache=FALSE}
xtable::xtable(allen_priors_xtable,
  caption = "Parameterization range for the uniform priors in the Allen model")
```

@@

@@

```{r echo=FALSE, include=TRUE, fig.cap = "Traces from the MCMC estimates of the GP model show reasonable mixing (no trend) and sampling rejection rate (no piecewise jumps)"}
gp_assessment_plots$traces_plot
```

@@

\newpage

@@

For simplicity we assume profit function $\Pi_t(X_t, h_t) is linear in the realized harvest, only
enforcing the restriction that harvest can not exceed available stock, that is $\Pi(h,x) = min(h,x)$. $Z_t$ is as previously defined.

@@

Where $V_t$ is the discounted value of the profit given that $X(t) = x$

@@

All code used in producing this analysis can be found in the R package accompanying the paper, available from [http://github.com/cboettig/nonparametric-bayes]. The code for all results shown here is dynamically embedded into the manuscript using `knitr` [@knitr].
